---
title: "Gapminder Analysis"
author: "Chris Toole & Max Pohlman"
date: "February 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Gapminder dataset analysis and trends.

```{r load data, echo=FALSE}
library(curl)
library(ggplot2)
gapminder_location<-curl(url="https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder<-read.csv(gapminder_location,stringsAsFactors = FALSE)

```

## Analysis
The first thing we did was we looked at the correlation between life expectancy and GDP per capita using the following code:

```{r firsttable, echo=TRUE, eval=FALSE}
cor(gapminder$lifeExp, gapminder$gdpPercap)
```
The correlation between life expextancy and GDP per capita is `r cor(gapminder$lifeExp, gapminder$gdpPercap)`

We then decided to visualize the relationship between the two variables by using a scatter plot.

```{r plotting, echo=FALSE}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

```{r pressure, echo=FALSE}
#plot lifeexp (y) vs time (x)
ggplot(data=gapminder, aes(x=year, y=lifeExp, color=continent))+
  geom_point()+geom_smooth(method=lm)
```
Above plot shows trend.