---
title: "Gapminder Analysis"
author: "Chris Toole & Max Pohlman"
date: "February 28, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Gapminder dataset analysis and trends.

```{r load data, echo=FALSE, message=FALSE}
library(magrittr)
library(ggplot2)
library(plyr)
library(dplyr)
library(curl)
gapminder_location<-curl(url="https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder<-read.csv(gapminder_location,stringsAsFactors = FALSE)
#function calculates total gdp
calcgdp <- function(dat,year=NULL,country=NULL){
  if(!is.null(year)) {
    dat<- dat[dat$year %in% year,]
  }
  if(!is.null(country)) {
    dat<- dat[dat$country %in% country,]
  }
  gdp<-dat$gdpPercap*dat$pop
  dat<-cbind(dat,gdp)
  return(dat)
}
newgap<-calcgdp(gapminder)

```

## Analysis
The first thing we did was we looked at the correlation between life expectancy and GDP per capita using the following code:

```{r firsttable, echo=TRUE, eval=FALSE}
cor(gapminder$lifeExp, gapminder$gdpPercap)
```
The correlation between life expextancy and GDP per capita is `r cor(gapminder$lifeExp, gapminder$gdpPercap)`

We then decided to visualize the relationship between the two variables by using a scatter plot.

```{r plotting, echo=FALSE, fig.align='center'}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()+ggtitle('Life exptectency vs gdp per capita') +
  theme(plot.title = element_text(hjust = 0.5))
```
This looks like there's a clear relationship between gdp and life expectancy! The plot appears exponential so the same plot was created below with a logarithmic scale of base 10 applied to the y-axis and then fit to a linear model. 

``` {r plottwo, echo=FALSE, fig.align='center', fig.width = 7, fig.height = 3}
obj<-lm(gdpPercap~lifeExp, data=gapminder)
ggplot(data=gapminder, aes(x=lifeExp, y=gdpPercap))+
  geom_point()+scale_y_log10()+geom_smooth(method=lm)+ theme(legend.position="none") +
  geom_text(aes(x=40, y=30000, label=(paste('The r-squared is ' , round(summary(obj)$r.squared,digits=3))),size=3)) #adds r squared
```
The above analysis was then performed on total GDP. The correlation between life expextancy and total GDP is `r cor(newgap$lifeExp, newgap$gdp)`. 


```{r plotting3, echo=FALSE, fig.align='center'}
ggplot(data = newgap, aes(x = lifeExp, y = gdp)) +
  geom_point()+ggtitle('Life exptectency vs total gdp') +
  theme(plot.title = element_text(hjust = 0.5))
```
It appears there are several exponential trends above, so the plot was then separated by continent.

```{r plotting4, echo=FALSE, fig.align='center', fig.width = 7, fig.height = 4}
ggplot(data = newgap, aes(x = lifeExp, y = gdp, color=continent)) +
  geom_point()+ggtitle('Life exptectency vs total gdp') +
  theme(plot.title = element_text(hjust = 0.5))
```
Then a logarithmic scale was applied to the y axis and the data for each continent was fit to a linear model but none appear to be highly correlated.

``` {r plot5, echo=FALSE, fig.align='center'}
ggplot(data=newgap, aes(x=lifeExp, y=gdp, color=continent))+
  geom_point()+scale_y_log10()+geom_smooth(method=lm)
```

Now lets take a look at mean life expectancy and gdpPercap by continent.
```{r barchart, echo=FALSE, fig.align='center', fig.width = 7, fig.height = 4}
plottable<-ddply(gapminder,~continent,summarise,lifeexp=mean(lifeExp)) 
ggplot(data = plottable, aes(continent, lifeexp, fill=continent)) +
   geom_bar(stat="identity")+ggtitle('Mean life expectancy per continent') +
  theme(plot.title = element_text(hjust = 0.5))
```

``` {r plot55, echo=FALSE, fig.align='center'}
gapminder %>% group_by(continent) %>% summarize(gdp=mean(gdpPercap)) %>%
  ggplot(aes(continent, gdp, fill=continent)) +
   geom_bar(stat="identity")+ggtitle('Mean gdp per capita by continent') +
  theme(plot.title = element_text(hjust = 0.5))
```
Further, these can be separated by year as well. 

```{r plotting101, echo=FALSE, fig.align='center', fig.width = 7, fig.height = 4}
plottable<-ddply(gapminder,c("continent",'year'),summarise, mle=mean(lifeExp)) #gets the mean for every continent in every year
pplottable<-ddply(plottable,c("continent"),summarise, up=max(mle), down=min(mle)) #gets the highest and lowest mean for each continent
ggplot()+
  geom_point(data=plottable, aes(x=continent,y=mle,color=year), size=3) + 
  geom_linerange(data=pplottable, aes(continent,ymin=down,ymax=up), 
                 color=c('orange','orange','orange','orange','orange'), 
                 size=c(1,1,1,1,1)) +
  geom_text(data=pplottable,  #Puts the numbers next to the lines
                 aes(continent, y=(pplottable$up+pplottable$down)/2,label=(round((pplottable$up-pplottable$down),digits=1))), hjust=1.5) +
  labs(y='Mean Life Expectancy')+
  ggtitle('Changes in Life Expectancy from 1952 to 2007 by Continent') +
  theme(plot.title = element_text(hjust = 0.5))

```
```{r plot66, echo=FALSE, fig.align='center',fig.width = 7.5}
plottable<-ddply(gapminder,c("continent",'year'),summarise, gdp=mean(gdpPercap)) #gets the mean for every continent in every year
pplottable<-ddply(plottable,c("continent"),summarise, up=max(gdp), down=min(gdp))#gets the highest and lowest mean for each continent
ggplot()+
  geom_point(data=plottable, aes(x=continent,y=gdp,color=year), size=3) + 
  geom_linerange(data=pplottable, aes(continent,ymin=down,ymax=up), 
                color=c('orange','orange','orange','orange','orange'), 
                size=c(1,1,1,1,1)) +
  geom_text(data=pplottable, #Puts the numbers next to the lines
            aes(continent, y=(pplottable$up+pplottable$down)/2,
                label=format((round((pplottable$up-pplottable$down),digits=1)),big.mark=',', scientific = FALSE), hjust=1.25)) +
  labs(y='Mean GDP')+
  ggtitle('Changes in GDP from 1952 to 2007 by Continent') +
  theme(plot.title = element_text(hjust = 0.5))

```
It appears the life expectancy of Asia has increased the most over time, while the life expectancy in Oceania has changed the least, with the highest overall mean. Oceania and Europe appear to have sustained approximately the same economic growth and have the highest change in gdp among all continents. Asia and the Americas also appear to have similar increases in gdp, ableit of less magnitude. Africa has the smallest increase in gdp as well as the lowest mean gdp in each year.


Unsurprisingly, Africa, the poorest continent, has the lowest life expectancy. However, this might have improved over time. We then plotted life expectancy over time for each continent.

```{r maxlinechart, echo=FALSE, message=FALSE, fig.align='center'}
asdf<-ddply(gapminder,c("continent","year"),summarise, lifeexp=mean(lifeExp)) 
ggplot(asdf, aes(year, lifeexp, color=continent)) +
  geom_smooth(se=FALSE)
```

It looks like Africa's average life expectancy increased at approximately the same rate as the other continents until around 1987.

Furthermore, the Americas seem low. This might be because North and South America are combined. Let's now split them up:

```{r americasone, echo=FALSE}
gapminder$america[gapminder$country=='Canada'|gapminder$country=='United States'] <- "North America"
gapminder$america[
  gapminder$country=='Mexico'|
  gapminder$country=='Costa Rica'|
  gapminder$country=='Cuba'|
  gapminder$country=='Dominican Republic'|
  gapminder$country=='Jamaica'|
  gapminder$country=='Puerto Rico'|
  gapminder$country=='Haiti'|
  gapminder$country=='Honduras'|
  gapminder$country=='Panama'|
  gapminder$country=='El Salvador'|
  gapminder$country=='Guatemala'|
  gapminder$country=='Nicaragua'
  ]<- "Central America"
gapminder$america[
    gapminder$country=='Brazil'|
    gapminder$country=='Argentina'|
    gapminder$country=='Bolivia'|
    gapminder$country=='Chile'|
    gapminder$country=='Colombia'|
    gapminder$country=='Ecuador'|
    gapminder$country=='Paraguay'|
    gapminder$country=='Peru'|
    gapminder$country=='Venezuela'|
    gapminder$country=='Trinidad and Tobago'|
    gapminder$country=='Uruguay'
  ]<- "South America"
NAnames<-unique(gapminder$country[gapminder$america=='North America'])
NAnames<-NAnames[!is.na(NAnames)]
CAnames<-unique(gapminder$country[gapminder$america=='Central America'])
CAnames<-CAnames[!is.na(CAnames)]
SAnames<-unique(gapminder$country[gapminder$america=='South America'])
SAnames<-SAnames[!is.na(SAnames)]
```
The countries we classified as North America were `r NAnames`

The countries we classified as Central America were `r CAnames`

The countries we classified as South America were `r SAnames`

```{r maxlinecharttwo, echo=FALSE, message=FALSE, fig.width = 7, fig.height = 4, fig.align='center'}
plottable<-ddply(subset(gapminder, continent == 'Americas'),c("america","year"),summarise, lifeexp=mean(lifeExp)) 
ggplot(plottable, aes(year, lifeexp, color=america)) +
  geom_smooth(se=FALSE)
```
Wow! There appears to be quite the difference between North America and the other two Americas. Let's verify this with a t-test for difference of means to see if they are significantly different.

```{r maxttest, echo=FALSE}
NAvalues<-gapminder$lifeExp[gapminder$america=='North America']
CASAvalues<-gapminder$lifeExp[gapminder$america=='Central America' | gapminder$america=="South America"]
t.test(NAvalues,CASAvalues)
```

With a t stat of a whopping 11.14, there is a significant difference between life expectancy in North America and the other two Americas.

Furthermore, after receiving peer review, it was suggested we include a paired t-test with NA and CA/SA paired on each year.
```{r maxttest2, echo=FALSE}
tester<-ddply(subset(gapminder, gapminder$america=='North America'),'year',summarise, na=mean(lifeExp))
testertwo<-ddply(subset(gapminder, gapminder$america=='Central America' | gapminder$america=='South America'),'year',summarise, casa=mean(lifeExp))
tester$casa<-testertwo$casa
t.test(tester$na,tester$casa, paired = TRUE)
```

Still a highly significant number suggesting a strong difference between North America and Central/South America life expectancies.

Lastly, population was plotted against life expectancy and gdp per capita was plotted against population. Neither of these plots displayed clear relationships.

```{r plotting_pop_lifeExp, echo=FALSE, fig.height=4, fig.align='center'}
ggplot(data = gapminder, aes(x = lifeExp, y = pop, color=continent)) +
  geom_point()+geom_smooth(method=lm)+ggtitle('Life exptectency vs population') +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r plotting_pop_gdp, echo=FALSE, fig.height=4, fig.align='center'}
ggplot(data = gapminder, aes(x = pop, y = gdpPercap, color=continent)) +
  geom_point()+ggtitle('Population vs gdp per capita') +
  theme(plot.title = element_text(hjust = 0.5))
```
